---
title: NOAA Storm Database exploring for US healt and economic consequences due to
  severe weather events
author: "Corrado Lanera"
date: "17 giugno 2015"
output: 
  html_document: 
    keep_md: yes
---

## Synopsis

> Most 10 complete sentences.


## Data Processing

### Preamble

To perform the analysis we use __R__ with the `package: dplyr`. Full code to provide the analysis are reported below whit its own explicit output. The code to produce the document can be found in mine Git repository [4].

```{r}
library(dplyr)                      # efficient tool for working with data frame
```

Final confifuration of the system is:

```{r}
sessionInfo()
```



### Loading \& processing the Raw Data

> Describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the cache = TRUE option for certain code chunks.

From __NOAA Storm Database__ [1] we obtain data on storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. We obtain file from 1950 up to November 2011. Some description of the data can be found in [2] and [3].

Data are stored in standard *comma-separated-value* format, compress with *bzip2* algorithm.

```{r eval=FALSE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
              "storm.data.bz2")                              # download raw data
```
```{r cache=TRUE}
storm.data <- read.csv(                                      # read csv from the
                bzfile("storm.data.bz2", "r"))            # compressed .bz2 file
```

### Reading \& selecting data

After loading data we have created the table to visualize information in a useful way (thanks to `package: dplyr`). 
Next we convert names in compatibility names both for R and for applications which do not allow *underline* in names.

```{r cache=TRUE}
storm <- tbl_df(storm.data)                       # create useful table for data

names(storm) <- make.names(names(storm), allow_ = FALSE)   # compatibility names

storm                                                     # read the data tables
```

There were `r dim(storm)[1]` observations with `r dim(storm)[2]` variables.

The variables we are interested in are the __type of event__ (`EVTYPE`), __fatalities__ (`FATALITIES`) and __injuries__ (`INJURIES`) and __ammount of damage__ (all fields including `DMG`). Here we extract those variables and print a sample of ten cases.

```{r}
use_storm <- storm %>%                                              # from strom
        select(EVTYPE, FATALITIES, INJURIES,         # select explicit variables
               contains("DMG"))                  # and the ones containing "DMG"

set.seed(1304)                  # set seed for reproducibility random selections
sample_n(use_storm,10)
```


## Results


> At least one figure containing a plot.

> Figures may have multiple plots in them (i.e. panel plots), but there cannot be more than three figures total.


### Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

In order to find the most harmful event with respect to population healt we firstly select only those variable concerning the type of event, fatalities and injuries. Next we consider the total ammount of both wrt the type of event. Next we consider the sum of fatalities and injuries and rank each events according to the number of fatalities, injuries and the sum of them. Finally we show the first ten events wrt to rank of total injuries and fatalities.

```{r cache=TRUE}
health_storm <- use_storm %>%                                   # from use_storm
                select(EVTYPE, FATALITIES, INJURIES) %>%  # select explicit vars
                group_by(EVTYPE) %>%         # grouped them by the type of event
                summarise_each(funs(sum)) %>% # compute the sum wrt to the group
                mutate(TOT.HARMFUL=FATALITIES + INJURIES, # create new variables
                       RK.FAT=dense_rank(desc(FATALITIES)),
                       RK.INJ=dense_rank(desc(INJURIES)),
                       RK.TOT=dense_rank(desc(TOT.HARMFUL)) 
                       ) %>%
                arrange(desc(TOT.HARMFUL),                 # arrange the dataset
                        desc(FATALITIES),
                        desc(INJURIES)
                        )

health_storm                                                         # show data

barplot(t(as.matrix(health_storm[1:10,4:2])),          # plot the first 10 cases
        main = "First 10 most harmful events wrt polulation healt",      # title
        names.arg = health_storm$EVTYPE[1:10],     # names of the groups of bars
        las=3,                         # 90 degrees rotation only for bar lables
        cex.names = 0.45,            # magnification of lables to fit the screen
        ylab = "People death/injuried",                               # y lables
        beside = TRUE,       # grouped bars (not in a single cumulative columns)
        col = terrain.colors(3))                                  # some colours
legend(20,50000,c("Total", "Injuried", "Fatalities"),                   # legend
       fill = terrain.colors(3))           # colours of the legend: same as bars
```

It is clear that the in all considered cases (only fatalities, only injuries or the total sum of them) __tornado is the most harmful wrt population healt__.

### Across the United States, which types of events have the greatest economic consequences?

As above, in order to find the events with the greatest economic consequences we firstly select only those variable concerning the type of event, and the one which report the damages ammount.


Next we consider the total ammount of both wrt the type of event. Next we consider the sum of fatalities and injuries and rank each events according to the number of fatalities, injuries and the sum of them. Finally we show the first ten events wrt to rank of total injuries and fatalities. 

```{r cache=TRUE}
PROP_storm <- use_storm %>%                                     # from use_storm
                select(EVTYPE, starts_with("PROP")) %>%   # select explicit vars
                group_by(EVTYPE, PROPDMGEXP) %>%    # grop by type and magnitude
                summarize(DAMAGE_SET=sum(PROPDMG)) %>%# sum wrt to the magnitude
                mutate(                     # compute explicit ammount of damage
                        DAMAGE=ifelse(PROPDMGEXP=="K",                  # K=10^3
                                     DAMAGE_SET*(10^3),
                                ifelse(PROPDMGEXP=="M",                 # M=10^6
                                       DAMAGE_SET*(10^6),
                                ifelse(PROPDMGEXP=="B",                 # B=10^9
                                       DAMAGE_SET*(10^9),
                                DAMAGE_SET))              # no changes otherwise
                                )
                       ) %>%
                summarise(TOTPROPDMG=sum(DAMAGE))     # comput the total ammount

CROP_storm <- use_storm %>%                                   # similar as above
                select(EVTYPE, starts_with("CROP")) %>%
                group_by(EVTYPE, CROPDMGEXP) %>% 
                summarize(DAMAGE_SET=sum(CROPDMG)) %>% 
                mutate(DAMAGE=ifelse(CROPDMGEXP=="K",
                                     DAMAGE_SET*(10^3),
                                ifelse(CROPDMGEXP=="M",
                                       DAMAGE_SET*(10^6),
                                ifelse(CROPDMGEXP=="B",
                                       DAMAGE_SET*(10^9),
                                DAMAGE_SET))
                                )
                       ) %>%
                summarise(TOTCROPDMG=sum(DAMAGE))

DMG_storm <- full_join(PROP_storm,CROP_storm) %>%                # join datasets
                mutate(TOTDMG=TOTPROPDMG + TOTCROPDMG,     # comput total damage
                       RK.PROP=dense_rank(desc(TOTPROPDMG)),         # and ranks
                       RK.CROP=dense_rank(desc(TOTCROPDMG)),
                       RK.DMG=dense_rank(desc(TOTDMG))) %>%
                arrange(desc(TOTDMG))            # arrange wrt the global damage

DMG_storm                                                            # show data

barplot(t(as.matrix(DMG_storm[1:10,c(4,2,3)])),          # as the previusly plot
        main = "First 10 events with greatest economic consequences",
        names.arg = DMG_storm$EVTYPE[1:10],
        las=3,
        cex.names = 0.45,
        ylab = "Damage in US $",
        beside = TRUE,
        col = terrain.colors(3))
legend(20,80000000000,c("Total", "Property", "Crop"),
       fill = terrain.colors(3))
```

As in the case of population healt it is one events that clearly is the worst one. In this case it is not the tornado (which reach the 3^rd^ position) but the _Flood is the events with the greatest economic consequences_.

## References

[1] Storm Data [47Mb]

https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

[2] National Weather Service Storm Data Documentation.

https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf

[3] National Climatic Data Center Storm Events FAQ.

https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf

[4] This assessment Git repo.

https://github.com/CorradoL/RepData_PeerAssessment2