---
title: NOAA Storm Database analysis for U.S. healt and economic consequences due to
  severe weather events
author: "Corrado Lanera"
date: "17 giugno 2015"
output:
  html_document:
    keep_md: yes
  pdf_document: default
  word_document: default
---

## Synopsis

> Most 10 complete sentences.

In this report we aim to find the sever weather events that had the worst consequences on population healt and economic in the United State from 1950 to 2011. We based our analysis on the National Oceanic and Atmospheric Administration (NOAA) Storm Database. The results of the analysis clearly identify the tornado as the most harmful event wrt population healt and the flood as the one with the greatest economic consequences. To perform the research we used the statistical program language R.

## Data Processing

### Preamble

To perform the analysis we use __R__ with the `package: dplyr`. Full code are reported below whit each explicit output. Morover the code to produce the present document is written in __R Markdown__ and can be found inside mine __GitHub__ repository [4].

```{r}
library(dplyr)                      # efficient tool for working with data frame
```

Final configuration of the system is:

```{r}
sessionInfo()         # provide principal info about configuration of the system
```



### Loading \& processing the Raw Data

> Describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the cache = TRUE option for certain code chunks.

From __NOAA Storm Database__ [1] we obtain data on storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. We obtain file from 1950 up to November 2011. Some description of the data can be found in [2] and [3].

Data are stored in the standard *comma-separated-value* format, compress with *bzip2* algorithm.

```{r cache=TRUE}
download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
              "storm.data.bz2")                              # download raw data

storm.data <- read.csv(                                      # read csv from the
                bzfile("storm.data.bz2", "r"))            # compressed .bz2 file
```

### Reading \& selecting data

After loading data we have created the table to visualize information in a useful way (thanks to `package: dplyr`). 
Next we convert names in *compatibility names* both for R and for applications which do not allow *underline* in names.

```{r cache=TRUE}
storm <- tbl_df(storm.data)                       # create useful table for data

names(storm) <- make.names(names(storm), allow_ = FALSE)   # compatibility names

storm                                                     # read the data tables
```

There were `r dim(storm)[1]` total observations with `r dim(storm)[2]` variables.

The variables we are interested in are the __type of event__ (`EVTYPE`), __fatalities__ (`FATALITIES`) and __injuries__ (`INJURIES`) and those describing the __ammount of damage__ (all fields including `DMG`). Here we extract those variables and print a sample of ten cases to watch them togheter.

```{r}
use.storm <- storm %>%                                              # from strom
        select(EVTYPE, FATALITIES, INJURIES,         # select explicit variables
               contains("DMG"))                  # and the ones containing "DMG"

set.seed(1304)                  # set seed for reproducibility random selections
sample_n(use.storm,10)                        # print a random sample of 10 rows
```


## Results


> At least one figure containing a plot.

> Figures may have multiple plots in them (i.e. panel plots), but there cannot be more than three figures total.


### Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

In order to find the most harmful event with respect to population healt we firstly select only those variable concerning the type of event, fatalities and injuries. Next we consider the total ammount of both wrt the type of event. Next we consider the sum of fatalities and injuries and rank each events according to the number of fatalities, injuries and the sum of them. Finally we show the first ten events wrt to rank of total injuries and fatalities.

```{r cache=TRUE}
health.storm <- use.storm %>%                                   # from use_storm
                select(EVTYPE, FATALITIES, INJURIES) %>%  # select explicit vars
                group_by(EVTYPE) %>%         # grouped them by the type of event
                summarise_each(funs(sum)) %>% # compute the sum wrt to the group
                mutate(TOT.HARMFUL=FATALITIES + INJURIES, # create new variables
                       RK.FAT=dense_rank(desc(FATALITIES)),
                       RK.INJ=dense_rank(desc(INJURIES)),
                       RK.TOT=dense_rank(desc(TOT.HARMFUL))) %>%
                arrange(desc(TOT.HARMFUL),                 # arrange the dataset
                        desc(FATALITIES),
                        desc(INJURIES))

health.storm                                                         # show data

barplot(t(as.matrix(health.storm[1:10,4:2])),# 4 interest vars of first 10 cases
        main = "First 10 most harmful events wrt polulation healt",      # title
        names.arg = health.storm$EVTYPE[1:10],     # names of the groups of bars
        las=3,                         # 90 degrees rotation only for bar lables
        cex.names = 0.45,            # magnification of lables to fit the screen
        ylab = "People death/injuried",                               # y lables
        beside = TRUE,       # grouped bars (not in a single cumulative columns)
        col = terrain.colors(3))                                  # some colours
legend(20,50000,c("Total", "Injuried", "Fatalities"),                   # legend
       fill = terrain.colors(3))           # colours of the legend: same as bars
```

It is clear that in all considered cases (only fatalities, only injuries or the sum of both) __tornado is the most harmful event wrt population healt__.

### Across the United States, which types of events have the greatest economic consequences?

As above, in order to find the event with the greatest economic consequences we firstly select only the variable of the dataset concerning the type of event, and the ones which report the damages ammount. More in detail we compute separately the property damage and the crop damage computing the explicit ammount of damage wrt the events. Next we rank the events wrt each of type of damage and for the total ammount of them. Finally we show the first ten events wrt to rank of total economic consequences.

```{r cache=TRUE}
PROP.storm <- use.storm %>%                                     # from use_storm
                select(EVTYPE, starts_with("PROP")) %>%   # select explicit vars
                group_by(EVTYPE, PROPDMGEXP) %>%    # grop by type and magnitude
                summarize(DAMAGE.SET=sum(PROPDMG)) %>%# sum wrt to the magnitude
                mutate(                     # compute explicit ammount of damage
                        PROPDAMAGE=ifelse(PROPDMGEXP=="K",              # K=10^3
                                     DAMAGE.SET*(10^3),
                                ifelse(PROPDMGEXP=="M",                 # M=10^6
                                       DAMAGE.SET*(10^6),
                                ifelse(PROPDMGEXP=="B",                 # B=10^9
                                       DAMAGE.SET*(10^9),
                                DAMAGE.SET)))) %>%        # no changes otherwise
                summarise(TOTPROPDMG=sum(PROPDAMAGE)) # comput the total ammount

CROP.storm <- use.storm %>%                                   # similar as above
                select(EVTYPE, starts_with("CROP")) %>%
                group_by(EVTYPE, CROPDMGEXP) %>% 
                summarize(DAMAGE.SET=sum(CROPDMG)) %>% 
                mutate(CROPDAMAGE=ifelse(CROPDMGEXP=="K",
                                     DAMAGE.SET*(10^3),
                                ifelse(CROPDMGEXP=="M",
                                       DAMAGE.SET*(10^6),
                                ifelse(CROPDMGEXP=="B",
                                       DAMAGE.SET*(10^9),
                                DAMAGE.SET)))) %>%
                summarise(TOTCROPDMG=sum(CROPDAMAGE))

DMG.storm <- full_join(PROP.storm,CROP.storm) %>%                # join datasets
                mutate(TOTDMG=TOTPROPDMG + TOTCROPDMG,     # comput total damage
                       RK.PROP=dense_rank(desc(TOTPROPDMG)),         # and ranks
                       RK.CROP=dense_rank(desc(TOTCROPDMG)),
                       RK.DMG=dense_rank(desc(TOTDMG))) %>%
                arrange(desc(TOTDMG))            # arrange wrt the global damage

DMG.storm %>% print(width = Inf)                       # show data with all vars

barplot(t(as.matrix(DMG.storm[1:10,c(4,2,3)])),          # as the previusly plot
        main = "First 10 events with greatest economic consequences",
        names.arg = DMG.storm$EVTYPE[1:10],
        las=3,
        cex.names = 0.45,
        ylab = "Damage in US $",
        beside = TRUE,
        col = terrain.colors(3))
legend(20,80000000000,c("Total", "Property", "Crop"),
       fill = terrain.colors(3))
```

As in the case of population healt it is one events that clearly is the worst one. In this case it is not the tornado (which reach the 3^rd^ position) but the _Flood is the events with the greatest economic consequences_.

## References

[1] Storm Data [47Mb]

https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

[2] National Weather Service Storm Data Documentation.

https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf

[3] National Climatic Data Center Storm Events FAQ.

https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf

[4] This assessment Git repo.

https://github.com/CorradoL/RepData_PeerAssessment2